### 2022-07-28 paper/localpsf.tex
paper edits to abstract and intro

% REQUIRED
\begin{abstract}
	We present a fast matrix-free method for approximating operators that have locally supported non-negative integral kernels. Such operators arise, for example, as Schur complements in Schur complement methods for solving partial differential equations (PDEs), Poincare-Steklov operators in domain decomposition methods, covariance operators in spatial statistics, blurring operators in imaging, and Hessians in distributed parameter PDE-constrained inverse problems. Our method computes impulse responses of the operator at a collection of scattered points, then interpolates these known impulse responses to evaluate entries of the integral kernel at other points where the impulse response is not known. 
%	Classical H-matrix construction methods require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products. 
	Impulse responses are computed by applying the operator to a small number of Dirac combs associated with ``batches'' of point sources. By solving an ellipsoid packing problem, we choose as many points as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. 
	The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Fast H-matrix methods are used to perform further matrix computations including matrix-matrix addition and multiplication, matrix inversion and factorization, and matrix-vector products in nearly-linear complexity.
	We apply the method to approximate the Hessian in a large-scale PDE-constrained inverse problem with highly informative data. Numerical results demonstrate that our method substantially outperforms existing methods. We are able to approximate high rank operators using only a small number of operator applies.
\end{abstract}

@@

We present a fast \emph{matrix-free} method for approximating operators, $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$, that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, $\gdim$ is the spatial dimension, and $L^2(\Omega)'$ is the space of continuous linear functionals on $L^2(\Omega)$ (i.e., the topological dual space). Such operators arise as Schur complements in Schur complements methods for solving partial differential equations (PDEs) \cite{le1997non,saad1999distributed,TOBYPAPER}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps) \cite{DTNMAP}, covariance operators in spatial statistics \cite{GeogaEtAl20,ChenStein21,LindgrenRueLindstrom11}, blurring operators in imaging \cite{DenisEtAl11,NagyOleary98}, and Hessians in distributed parameter PDE-constrained optimization and inverse problems \cite{AlgerEtAl19,IsaacEtAl15}.

By ``matrix-free,'' we mean that we have a black-box computational procedure through which we may apply $\Aop$ and its transpose\footnote{Recall that the transpose $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop w\right)(u)$ for all $u,w \in L^2(\Omega)$, where the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for the analogous operations with $\Aop^T$. }, $\Aop^T$, to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
but we do not have any further access to $\Aop$. In particular, we cannot easily access entries of $\Aop$'s integral kernel. In applications, evaluating the maps in \eqref{eq:eval_maps} may involve iteratively solving a large linear system, timestepping, or performing some other costly computational procedure.

Our motivation is Hessian approximation in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field from noisy observations of a state variable that depends on the parameter implicitly through the solution of a PDE.
In the deterministic approach to the inverse problem, one solves a regularized minimization problem to find the parameter that best fits the data. The objective function in the minimization problem is defined implicitly, so information about the Hessian of the objective function is only accessible via application of the Hessian to vectors \cite{HESSIANACTION}. 
The most popular matrix-free Hessian approximation methods are based on low rank approximation \cite{BuiEtAl13,CuiEtAl14,FlathEtAl11,PetraEtAl14,SpantiniEtAl15} because there are efficient methods for constructing low rank approximations of a matrix using only matrix-vector products \cite{Cheng05,HalkoMartinssonTropp11}. 
However, low rank approximation methods suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator applies are required to form an accurate ``low rank'' approximation \cite{Alger19,AlgerEtAl17}. High rank approximation methods are needed.

@@

The central challenge we face is that existing scalable high rank approximation methods typically require the ability to rapidly evaluate arbitrary entries of $\Aop$'s integral kernel, and we cannot do this because $\Aop$ is only accessible matrix-free. We address this challenge using \emph{impulse response interpolation}. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(w) = \left(\rho^*,w\right)_{L^2(\Omega)}$ for all $w \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate translated versions of these impulse responses to approximate arbitrary entries of the operator's integral kernel. 
%This allows fast evaluation of approximate matrix entries, which are needed by high-rank approximation methods, in particular, classical H-matrix construction methods.

Picking the batches of points requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. The idea of estimating the supports of the functions $\phi_x$ a-priori was inspired by techniques from resolution analysis in seismic imaging. There, $\Aop^T$ is applied to a random noise function, and the width of $\phi_x$ is estimated to be the autocorrelation length of the resultant function near $x$  \cite{FichtnerLeeuwen15,TrampertFichtnerRitsema13}. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Section \ref{sec:intromoments}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing at the cost of the additional constraint that $\Aop$ has a positive integral kernel.

Our impulse response interpolation procedure may be categorized as a ``point spread function'' (PSF) method that is loosely based on so-called ``product-convolution'' (PC) approximations, which are approximations of an operator by weighted sums of convolution operators with spatially varying weights. PC and PSF methods have a long history dating back several decades. We note the following papers (among many others),  \cite{Adorf94,AlgerEtAl19,BigotEscandeWeiss19,NagyOleary98,FishEtAl96,EscandeWeiss12,EscandeWeiss15,EscandeWeiss22,ZhuLiFomelEtAl16}, in which the convolution kernels are constructed from sampling impulse responses of the operator to scattered point sources. For background on PC and PSF methods, we recommend the following papers which provide well-written reviews of the topic: \cite{DenisEtAl15,EscandeWeiss17,GentileCourbinMeylan13}. While PC approximations are based on an assumption of local translation invariance, the method we propose is based on an assumption that we call ``local mean displacement invariance'' (see Figure \ref{fig:mean_displacement_invariance}). This is more general than local translation invariance, and includes both low rank approximation and local translation invariance as special cases (see Section \ref{sec:local_mean_displacement_invariance}). Furthermore, PC approximations have well-known issues near boundaries due to so-called ``boundary artifacts,'' caused by the boundary artificially ``cutting off'' impulse responses. We avoid this issue by temporarily excluding impulse responses from the approximation procedure whenever an attempt is made to evaluate the impulse response at a location where it is undefined.

@@

The ability to rapidly approximate entries of $\Aop$'s integral kernel allows us to form a hierarchical matrix (H-matrix) \cite{Hackbusch99,} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high rank. 
%(see Figure \ref{fig:hmatrix_intro} and Appendix \ref{app:h_matrix}).
H-matrix methods permit us to perform matrix-vector products cheaply, and perform other useful linear algebra operations that cannot be done easily using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. 
%These H-matrix methods are fast and scale well to large problems. 
The work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix scales 
%as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. That is, the cost is \emph{nearly-linear}
\emph{nearly-linearly} in $N$ (i.e., $o(N^{1+\epsilon})$ for any $\epsilon > 0$). The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks; for more details see \cite{GrasedyckHackbusch03}. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

%Classical H-matrix construction methods 
%(described in Appendix \ref{sec:H_matrix_construction}) 
%require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available matrix free.
Recently there have been improvements in matrix-free H-matrix construction methods based on the so-called ``peeling process'' \cite{LinYing11,Martinsson11,Martinsson16,MartinssonTropp20}, which we do not use here. 
These alternative methods have been applied to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{AmbartsumyanEtAl20,HartlandEtAl22}. 
Methods based on the peeling process are asymptotically scalable (typically the number of matrix vector products required scales as $O(\log N)$), but in practice the required number of matrix vector products is large. Improving the peeling process to reduce its cost is currently an area of active research \cite{PEELINGRESEARCH}.
%This is an area of active research, so more efficient and practical matrix-free H-matrix construction methods based on the peeling process may be discovered in the future.
%Matrix-free H-matrix construction methods are currently a topic of considerable research, so more efficient matrix-free H-matrix methods based on the peeling process may be discovered in the future.

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.85]{mean_displacement_invariance.pdf}
	\end{center}
	\caption{An operator is locally mean-displacement invariant if $\impulseresponse_{x}(y) \approx \impulseresponse_{x'}\left(y - \spatialmean(x) + \spatialmean(x')\right)$ when $x$ is close to $x'$. Here $\spatialmean(x)$ is the mean (center of mass) of $\phi_x$, and similar for $\spatialmean(x')$.}
	\label{fig:mean_displacement_invariance}
\end{figure}

@@

The action of $\Aop$ may be extended to distributions if the kernel $\Aker$ is sufficiently regular. We derive the extension as follows. Suppose $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(w) &= \int_\Omega \int_\Omega w(y) \Aker(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega w(y) \int_\Omega \Aker(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega w(y) \genericdistribution\left(\Aker(y,\cdot)\right) dy, \label{eq:action_on_distribution}
	\end{align}
\end{subequations}
where $\Aker(y,\cdot)$ denotes the function $x \mapsto \Aker(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions, and instead suppose $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ is a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in \eqref{eq:extension_to_distributions} is not valid. However, if $\Aker$ is sufficiently regular such that the function $y \mapsto \rho\left(\Aker(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the right hand side of \eqref{eq:action_on_distribution} is well-defined. Hence, we \emph{define} the action of $\Aop$ on the distribution $\rho$ to be the right hand side of \eqref{eq:action_on_distribution}. For convenience of notation, we denote this action by ``$\Aop \genericdistribution^*$,'' even if $\rho^*$ does not exist.

@@

The direct approach to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ is to apply $\Aop$ to a point source centered at $x$ to obtain $\impulseresponse_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\impulseresponse_x$ to determine $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$. However, this direct approach is not feasible because we need to know $V(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ before we compute $\phi_x$, in order choose the point $x$. Computing $\phi_x$ in order to determine $V(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ would be extremely computationally expensive, and defeat the purpose of our algorithm, which is to reduce the computational cost by computing impulse responses in batches. Fortunately, it is possible to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ indirectly, \emph{for all points $x\in\Omega$ simultaneously}, by applying $\Aop^T$ to one constant function, $\gdim$ linear functions, and $\gdim(\gdim+1)/2$ quadratic functions (e.g., 6 total operator applies in two spatial dimensions and 10 in three spatial dimensions). This may be motivated by analogy to matrices. If $\mathbf{B}\in \mathbb{R}^{\fedim \times \fedim}$ is a matrix with $i^\text{th}$ column $\mathbf{b}_i$ and $\mathbf{f} \in \mathbb{R}^\fedim$, then
\begin{equation*}
\mathbf{B}^T \mathbf{w} = \begin{bmatrix}
\horzbar & \mathbf{b}_1^T & \horzbar \\
\horzbar & \mathbf{b}_2^T & \horzbar \\
& \vdots & \\
\horzbar & \mathbf{b}_\fedim^T & \horzbar
\end{bmatrix}
\mathbf{w} = 
\begin{bmatrix}
\mathbf{b}_1^T \mathbf{w} \\
\mathbf{b}_2^T \mathbf{w} \\
\vdots \\
\mathbf{b}_\fedim^T \mathbf{w}
\end{bmatrix}.
\end{equation*}
By computing one matrix-vector product of $\mathbf{B}^T$ with $\mathbf{w}$, we compute the inner product of each column of $\mathbf{B}$ with $\mathbf{w}$ simultaneously. The operator case is analogous, with the impulse response $\phi_x$ taking the place of a matrix column. We have
\begin{equation}
\label{eq:operator_simultaneous_ips}
\left(\Aop^T w\right)^*(x) = \int_\Omega \Aker(y,x) w(y) dy = \left(\phi_x, w\right)_{L^2(\Omega)}.
\end{equation}
By computing one operator application of $\Aop^T$ to $w$, we compute the inner product of each impulse response $\phi_x$ with $w$, for all points $x$ simultaneously.

>>>>

% REQUIRED
\begin{abstract}
	We present an efficient matrix-free point spread function (PSF) method for approximating operators that have locally supported non-negative integral kernels. Such operators arise as Schur complements in Schur complement methods for solving partial differential equations (PDEs), Poincare-Steklov operators in domain decomposition methods, covariance operators in spatial statistics, blurring operators in imaging, and Hessians in PDE-constrained inverse problems. We compute impulse responses of the operator at a collection of scattered points, then interpolate translated versions of these impulse responses to evaluate entries of the integral kernel at other points where the impulse response is not known. The interpolation uses a new approximation we call ``local mean displacement invariance,'' which improves upon classical local translation invariance.
%	Classical H-matrix construction methods require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products. 
	Impulse responses are computed by applying the operator to a small number of Dirac combs associated with ``batches'' of point sources. By solving an ellipsoid packing problem, we choose as many point sources as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. Support ellipsoids for the impulse responses are estimated a-priori by applying the operator to a small number of low order polynomial functions, then post-processing the resulting functions to determine impulse response moments for all points simultaneously.
	The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Fast H-matrix methods are used to perform further matrix computations including matrix-matrix addition and multiplication, matrix inversion and factorization, and matrix-vector products in nearly-linear complexity.
	We apply the method to approximate the Hessian in a large-scale PDE-constrained inverse problem with highly informative data. Numerical results demonstrate that our method substantially outperforms existing methods. We are able to approximate high rank operators using only a small number of operator applies.
\end{abstract}

@@

We present a fast \emph{matrix-free} point spread function (PSF) method for approximating operators, $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$, that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, $\gdim$ is the spatial dimension, and $L^2(\Omega)'$ is the space of continuous linear functionals on $L^2(\Omega)$ (i.e., the topological dual space). Such operators arise as Schur complements in Schur complements methods for solving partial differential equations (PDEs) \cite{le1997non,saad1999distributed,TOBYPAPER}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps) \cite{DTNMAP}, covariance operators in spatial statistics \cite{GeogaEtAl20,ChenStein21,LindgrenRueLindstrom11}, blurring operators in imaging \cite{DenisEtAl11,NagyOleary98}, and Hessians in distributed parameter PDE-constrained optimization and inverse problems \cite{AlgerEtAl19,IsaacEtAl15}. By ``matrix-free,'' we mean that we have a black-box computational procedure through which we may apply $\Aop$ and its transpose\footnote{Recall that the transpose $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop w\right)(u)$ for all $u,w \in L^2(\Omega)$, where the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for the analogous operations with $\Aop^T$. }, $\Aop^T$, to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
but we do not have further access to $\Aop$. In particular, we cannot easily access entries of $\Aop$'s integral kernel. In applications, evaluating the maps in \eqref{eq:eval_maps} may require performing a costly computational procedure, such as solving a large linear system.

Our motivation is Hessian approximation in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field from noisy observations of a state variable that depends on the parameter implicitly through the solution of a PDE.
In the deterministic approach to the inverse problem, one solves a regularized minimization problem to find the parameter that best fits the data. The objective function in the minimization problem is defined implicitly, so information about the Hessian of the objective function is only accessible via application of the Hessian to vectors \cite{HESSIANACTION}. 
The most popular matrix-free Hessian approximation methods are based on low rank approximation \cite{BuiEtAl13,CuiEtAl14,FlathEtAl11,PetraEtAl14,SpantiniEtAl15} because there are efficient methods for constructing low rank approximations of a matrix using only matrix-vector products \cite{Cheng05,HalkoMartinssonTropp11}. 
However, low rank approximation methods suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator applies are required to form an accurate ``low rank'' approximation \cite{Alger19,AlgerEtAl17}. Matrix-free high rank approximation methods are needed.

@@

%Existing scalable high rank approximation methods typically require the ability to rapidly evaluate arbitrary entries of $\Aop$'s integral kernel, which we cannot do. 
We form a high rank approximation using \emph{impulse response interpolation}. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(w) = \left(\rho^*,w\right)_{L^2(\Omega)}$ for all $w \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate translated versions of these impulse responses to approximate entries of the operator's integral kernel. 
%This allows fast evaluation of approximate matrix entries, which are needed by high-rank approximation methods, in particular, classical H-matrix construction methods. 
Picking the batches of points requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. The idea of estimating the supports of the functions $\phi_x$ a-priori was inspired by techniques from resolution analysis in seismic imaging. There, $\Aop^T$ is applied to a random noise function, and the width of $\phi_x$ is estimated to be the autocorrelation length of the resultant function near $x$  \cite{FichtnerLeeuwen15,TrampertFichtnerRitsema13}. We use polynomial functions instead of random noise functions (see Section \ref{sec:intromoments}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing at the cost of the additional constraint that $\Aop$ has a non-negative integral kernel. Our impulse response interpolation may be categorized as a PSF method that is loosely based on so-called ``product-convolution'' (PC) approximations, which are approximations of an operator by weighted sums of convolution operators with spatially varying weights. PC and PSF methods have a long history dating back several decades. We note the following papers (among many others),  \cite{Adorf94,AlgerEtAl19,BigotEscandeWeiss19,NagyOleary98,FishEtAl96,EscandeWeiss12,EscandeWeiss15,EscandeWeiss22,ZhuLiFomelEtAl16}, in which the convolution kernels are constructed from sampling impulse responses of the operator to scattered point sources. For background on PC and PSF methods, we recommend the following papers: \cite{DenisEtAl15,EscandeWeiss17,GentileCourbinMeylan13}. While PC approximations are based on an assumption of local translation invariance, the method we propose is based on an assumption that we call ``local mean displacement invariance,'' which is more general than local translation invariance, and includes both CUR low rank approximation and local translation invariance as special cases (see Section \ref{sec:local_mean_displacement_invariance} and Figure \ref{fig:mean_displacement_invariance}). 
%Furthermore, PC approximations have well-known issues near boundaries due to so-called ``boundary artifacts,'' caused by the boundary artificially ``cutting off'' impulse responses. We avoid this issue by temporarily excluding impulse responses from the approximation procedure whenever an attempt is made to evaluate the impulse response at a location where it is undefined.

@@

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.85]{mean_displacement_invariance.pdf}
	\end{center}
	\caption{$\Aop$ is locally mean-displacement invariant if $\impulseresponse_{x}(y) \approx \impulseresponse_{x'}\left(y - \spatialmean(x) + \spatialmean(x')\right)$ when $x$ is close to $x'$. Here $\spatialmean(z)$ denotes the mean (center of mass) of $\phi_z$.}
	\label{fig:mean_displacement_invariance}
\end{figure}

The ability to rapidly approximate entries of $\Aop$'s integral kernel allows us to form a hierarchical matrix (H-matrix) \cite{Hackbusch99,} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high rank. 
%(see Figure \ref{fig:hmatrix_intro} and Appendix \ref{app:h_matrix}).
H-matrix methods permit us to perform matrix-vector products cheaply, and perform other useful linear algebra operations that cannot be done easily using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. 
%These H-matrix methods are fast and scale well to large problems. 
The work and memory required to perform these operations for an $\fedim \times \fedim$ H-matrix scales 
%as $O(\fedim \log(\fedim)^a)$, where $a \in \{0,1,2,3\}$. That is, the cost is \emph{nearly-linear}
\emph{nearly-linearly} in $N$ (i.e., $o(N^{1+\epsilon})$ for any $\epsilon > 0$). The exact cost depends on the type of H-matrix used, the operation being performed, and the rank of the off-diagonal blocks (see \cite{GrasedyckHackbusch03}). The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification. 
%Classical H-matrix construction methods 
%(described in Appendix \ref{sec:H_matrix_construction}) 
%require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available matrix free.
Recently there have been improvements in matrix-free H-matrix construction methods based on the so-called ``peeling process'' \cite{LinYing11,Martinsson11,Martinsson16,MartinssonTropp20}, which we do not use. 
These alternative methods have been applied to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{AmbartsumyanEtAl20,HartlandEtAl22}. 
Methods based on the peeling process are asymptotically scalable,
%(typically the number of matrix vector products required scales as $O(\log N)$), 
but in practice the required number of matrix vector products is large. 
%Improving the peeling process to reduce its cost is currently an area of active research \cite{PEELINGRESEARCH}.
%This is an area of active research, so more efficient and practical matrix-free H-matrix construction methods based on the peeling process may be discovered in the future.
%Matrix-free H-matrix construction methods are currently a topic of considerable research, so more efficient matrix-free H-matrix methods based on the peeling process may be discovered in the future.

@@

The action of $\Aop$ may be extended to distributions if $\Aker$ is sufficiently regular. We derive the extension as follows. Suppose $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(w) &= \int_\Omega \int_\Omega w(y) \Aker(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega w(y) \int_\Omega \Aker(y,x) \genericdistribution^*(x) dx ~dy 
		= \int_\Omega w(y) \genericdistribution\left(\Aker(y,\cdot)\right) dy, \label{eq:action_on_distribution}
	\end{align}
\end{subequations}
where $\Aker(y,\cdot)$ denotes the function $x \mapsto \Aker(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions, and instead suppose $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ is a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in \eqref{eq:extension_to_distributions} is not valid. However, if $\Aker$ is sufficiently regular such that the function $y \mapsto \rho\left(\Aker(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the right hand side of \eqref{eq:action_on_distribution} is well-defined. Hence, we \emph{define} the action of $\Aop$ on the distribution $\rho$ to be the right hand side of \eqref{eq:action_on_distribution}. For convenience of notation, we denote this action by ``$\Aop \genericdistribution^*$,'' even if $\rho^*$ does not exist.

@@

The direct approach to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ is to apply $\Aop$ to a point source centered at $x$ to obtain $\impulseresponse_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\impulseresponse_x$ to determine $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$. However, this direct approach is not feasible because we need to know $V(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ before we compute $\phi_x$, in order choose the point $x$. Computing $\phi_x$ in order to determine $V(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ would be extremely computationally expensive, and defeat the purpose of our algorithm, which is to reduce the computational cost by computing impulse responses in batches. Fortunately, it is possible to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ indirectly, \emph{for all points $x\in\Omega$ simultaneously}, by applying $\Aop^T$ to one constant function, $\gdim$ linear functions, and $\gdim(\gdim+1)/2$ quadratic functions (e.g., 6 total operator applies in two spatial dimensions and 10 in three spatial dimensions). This may be motivated by analogy to matrices. If $\mathbf{B}\in \mathbb{R}^{\fedim \times \fedim}$ is a matrix with $i^\text{th}$ column $\mathbf{b}_i$ and $\mathbf{w} \in \mathbb{R}^\fedim$, then
\begin{equation*}
\mathbf{B}^T \mathbf{w} = \begin{bmatrix}
\horzbar & \mathbf{b}_1^T & \horzbar \\
\horzbar & \mathbf{b}_2^T & \horzbar \\
& \vdots & \\
\horzbar & \mathbf{b}_\fedim^T & \horzbar
\end{bmatrix}
\mathbf{w} = 
\begin{bmatrix}
\mathbf{b}_1^T \mathbf{w} \\
\mathbf{b}_2^T \mathbf{w} \\
\vdots \\
\mathbf{b}_\fedim^T \mathbf{w}
\end{bmatrix}.
\end{equation*}
By computing one matrix-vector product of $\mathbf{B}^T$ with $\mathbf{w}$, we compute the inner product of each column of $\mathbf{B}$ with $\mathbf{w}$ simultaneously. The operator case is analogous, with the impulse response $\phi_x$ taking the place of a matrix column. We have
\begin{equation}
\label{eq:operator_simultaneous_ips}
\left(\Aop^T w\right)^*(x) = \int_\Omega \Aker(y,x) w(y) dy = \left(\phi_x, w\right)_{L^2(\Omega)}.
\end{equation}
By computing one operator application of $\Aop^T$ to $w$, we compute the inner product of each impulse response $\phi_x$ with $w$, for all points $x$ simultaneously.
